On the properties of discs around accreting brown dwarfs

### 2010-05-18 BD_disc/paper.tex
more updates th
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2270 25464f21-6099-402c-b30a-640f1d40d38b

We have also applied dust sublimation during our simulations. This
treatment is similar to that detailed in \cite{2007ApJ...661..374T},
but with some enhancements to ensure convergence of the sublimated
rim.

The dust sublimation proceeds as follows: An initial temperature
distribution is found for the optically-thin limit by setting the
global dust-to-gas ratio to a tiny value. Subsequent radiative
equilibrium iterations are performed using the adopted dust-to-gas
ratio of 0.01, but limiting the maximum optical depth across a given
cell to $\tau_{\rm max}$. Cells whose temperature exceeds the local
dust sublimation temperature have their dust-to-gas ratio set to
zero. Radiative equilibrium iterations and sublimation sweeps are
performed at $\tau_{\max} = $0.1, 1, and 10, with a final iteration of
$\tau_{\max} = \infty$. Adequately solving the radiative-equilibrium
necessiates resolving the disc's effective photosphere, and adaptive
mesh refinement is used to split the grid at the
optically-thin/optically-thick boundary in order that the maximum cell
size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once a
self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations.

As our models are magnetically truncated at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range
\citep{2001ApJ...560..957D}. In fact the structure, radius and
generation mechanism of the inner wall will lead to markedly different
contributions to the SED. As discussed in Section \ref{intro} several
mechanisms can act to clear the inner disc.

Binarity and Giant planet formation would be expected to clear an
inner disc and shut down accretion from within the inner wall due to
gravitational perturbations. In addition, photoevaporation from UV and
FUV flux is expected to halt accretion from radii within the inner
disc due to the removal of this material via a stellar/disc wind
\citep{2006MNRAS.369..229A}. These two clearing mechanisms could be
differentiated by the resulting masses of the discs
\citep{2007MNRAS.378..369N}. However, for our model systems we have
neglected both of these effects.

For our models we have initially cleared the inner disc by invoking a
magnetic truncation radius, however, further disc clearing will be
caused by the dust destruction/sublimation mechanism we have included.
In some cases where the radius in the disc at which the dust
sublimation temperature and densities are reached, ($R_{\rm sub}$)
will be less than the co-rotation radius ($R_{\rm co}$). For these
cases, $R_{\rm sub}<R_{\rm co}$, the inner wall radius and shape will
be controlled by sublimation. Effectively meaning the final inner disc
radius, after the radiative transfer simulation, will not equal the
initial inner disc radius. The differences in the starting and
resulting inner wall radii and the shapes of the inner edges are
discussed in Section \ref{inner_edge}. If the inner disc is cleared
via dust sublimation one would expect a residual hot gas disc interior
to the inner wall. This is as the gas component will not be destroyed
and is free to extend to the co-rotation radius.  Evidence for a hot
gas inner disc component has been found through SED fitting by
\cite{2005ApJ...623..952E} and \cite{2008ApJ...689..513T}.  In
particular, \cite{2008ApJ...689..513T}, found that emission apparent
from within the inner wall radius could not be well fit by a modeled
spherical halo, or optically thin disc, dust component.

Furthermore, as the dust sublimation limit is temperature and density
dependent \citep{1994ApJ...421..615P}, dust sublimation can lead to
changes in shape of the inner wall. As the density of the
circumstellar disc will vary with scale height one would expect the
dust sublimation radius to vary concurrently with scale height. This
results in a curved inner wall \citep{2001A&A...371..186N}. As
explained by \cite{2007ApJ...661..374T}, the fact that near-IR
excesses of disc systems do not appear strongly dependent on
inclination indicates that the inner walls are not
vertical. \cite{2007ApJ...661..374T} also showed that grain growth and
midplane settling will lead to a curved inner wall (and a hot inner
gas disc). The curvature is caused by changes in the vertical dust
density distributions as a function of dust size, or midplane
settling. Essentially, larger grains will sink more efficiently
towards the midplane of the disc, and due to their larger surface
areas cool more efficiently. Therefore, as the scaleheight increases
the relative proportion of larger to smaller grains will reduce and
the cooling become less efficient. This will result in higher
temperatures being reached for larger scale heights at a fixed radius,
thereby curving the inner wall as the upper layers of the disc reach
the sublimation temperature at closer radii.
\cite{2007ApJ...661..374T}, modeled this effect by using two dust
distributions, a small and a large, with different vertical
distributions indicative of increased midplane settling for the larger
grains. For this study we have adopted a size distribution for one
population of dust grains, with one characteristic vertical
distribution. As such curvature of the inner wall in our simulations
will be dominated simply by the dust sublimation temperature, and
density. The resulting shapes and radial positions of the inner walls
after the radiative transfer simulations are complete are discussed in
detail in Section \ref{inner_edge}.

For our model grid we have adopted a gas-to-dust ratio, $\epsilon$, of
100, and the gas population is assumed essentially static with a zero
optical depth. Therefore, it does not contribute to the resulting SED,
and we have not included a hot inner gas disc as found by
\cite{2008ApJ...689..513T}.

@@

Uncertainty remains over the formation mechanisms, resulting IMFs and
subsequent interaction mechanisms with associated circumstellar disc,
for BDs. Furthermore, no current theoretical explanation can be
demonstrated for the suspected stellar mass-accretion rate
relationship \citep[$\dot{M}\propto
M_*^2$,][]{2003ApJ...592..266M,2004A&A...424..603N,2006A&A...452..245N}
derived from observations. In addition doubt remains over the
existence of a disc-locking mechanism in BDD systems, with predictions
of a correlation of rotation rate with infrared excess possibly
providing an intrinsic bias \citep{1997AJ....114..288M}. Theoretical
models are built upon constraints, such as ages, masses and disc
fractions, and correlations, such as the accretion rate to stellar
mass relation and rotation rate to period relation (disc-locking),
derived from the data. Critically, these resulting constraints and
relationships involve both the application of existing models, such as
isochrones to derive ages, and the selection of bias-minimised
samples, for instance removal of foreground or background contaminant
stars. Clearly, if one is to adopt a new theoretical model or use an
observed correlation as a constraint one must explore the suspected
intrinsic biases and the reliability of the selection criteria.

For our model grid we have selected a range of input parameters
covering expected, from observations, ranges. We have input the
fundamental parameters of age and mass by adoption of a set of
theoretical isochrones. We have intrinsically included a correlation
of rotation rate with initial inner edge position by setting this at
the co-rotation radius, which is expected to result in a correlation
of rotation rate with IR excess \citep{1997AJ....114..288M} (the
reverse of that expected due to angular momentum loss through
disc-locking). We have not assumed a relationship between accretion
rate and stellar mass for our model grid. Therefore, analysis of our
model grid in an attempt to recover the fundamental parameters of age
and mass, and show accretion rate to stellar mass or rotation rate to
infrared excess relationships, will allow us to assess the reliability
of these current theoretical models or relationships.

@@

The initial density structure in our BDD systems is setup as described
in \cite{2004MNRAS.351..607W} with the the disc surface density
conserved, $\Sigma ^{\beta - \alpha}$. The initial scaleheight,
$h\propto r^{\beta}$, where r is the radial distance from the star,
and the density, $\rho \propto r^{-\alpha}$. The initial values of
$\alpha$ and $\beta$ were 2.1 and 1.1 respectively \citep[see][for
full discussion of disc structure]{2004MNRAS.351..607W}. The values
for $\alpha$ and $\beta$ were chosen to optimise resolution of the
vertically evolving disc, but minor variations are largely
inconsequential as the systems evolves from this state. In our models
we have, however, placed at the inner disc edge at the co-rotation
radius, as opposed to the dust destruction radius used in
\cite{2004MNRAS.351..607W}. The initial disc scaleheight at 100 AU,
$h(100)$, was set to 25 AU. As the simulation used vertical
hydrostatic equilibrium and dust sublimation, both the disc
scaleheight and inner edge location then evolved in the systems
dependent on the input parameters. In this section we discuss the
structure of the discs in terms of these two generated
characteristics, i.e the disc scaleheight and inner edge location.

@@

As discussed in \cite{2004MNRAS.351..607W} discs around CTTS stars
have scaleheights at 100 AU of between $h$(100)$=$7 to 20 AU, whereas
for BDD systems, $h$(100)=20 to 60 AU (for 0.08 and 0.01 $M_{\odot}$
respectively). Our models adopt the same formalism for solving
hydrostatic equilibrium and therefore, at negligible accretion rates,
our disc scaleheights are comparable. As the accretion rate increases
the flux levels of the central star increase and lead to heating of
the disc. This in turn leads to vertical expansion in the disc. For
our models we found that levels of vertical flaring increased
marginally with accretion rate. Significant differences, more than
$>$5 AU increase in $h(50)$, in the vertical structure were not
apparent until the high accretion rates of \logmdot = $-7$ and
$-6$. Figures \ref{flare_-12_183} and \ref{flare_-7_193} show the
density structure ($log({\rho})$) in the disc from radial distances of
0 to 50 AU for example systems ($M_*=0.04M_{\odot}$, Age=1 Myrs,
$\tau$=5 and areal coverage=10\%), with accretion rates of \logmdot =
$-12$ and $-7$, respectively.

@@

\subsubsection{Inner Edge}
\label{inner_edge}

@@

\subsubsection{Accretion Dominance}
\label{acc_dominance}

@@

\subsubsection{Flaring and Obscuration}
\label{flare_obscure}

>>>>

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range
\citep{2001ApJ...560..957D}.

For low-luminosity systems the dust sublimation radius ($R_{\rm sub}$)
will be coincident with the co-rotation radius. However should
the combined photospheric and accretion luminosities be sufficient the
inner disc will be heated sufficiently that the dust close to the
co-rotation radius will be sublimated, and we must account for this
in our models. Furthermore, since the dust sublimation temperature has
a density dependence \citep{1994ApJ...421..615P}, it is both the location and {\em
  shape} of the inner wall that can change (Isella ref; Tannirkulam ref).

Our treatment of dust sublimation is similar to that detailed in
\cite{2007ApJ...661..374T}, but with some enhancements to ensure a
swift convergence of the sublimated rim. The dust sublimation proceeds
as follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessiates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations.

For our model grid we have adopted a gas-to-dust ratio, $\epsilon$, of
100, and the gas population is assumed essentially static with a zero
optical depth.

@@

@@

The initial density structure has the the disc surface density
conserved, $\Sigma ^{\beta - \alpha}$. The initial scaleheight,
$h\propto r^{\beta}$, where r is the radial distance from the star,
and the density, $\rho \propto r^{-\alpha}$. The initial values of
$\alpha$ and $\beta$ were 2.1 and 1.1 respectively \citep[see][for
  full discussion of disc structure]{2004MNRAS.351..607W}. The values
for $\alpha$ and $\beta$ were chosen to optimise resolution of the
vertically evolving disc, but minor variations are largely
inconsequential as the systems evolves from this state. In our models
we have, however, placed at the inner disc edge at the co-rotation
radius, as opposed to the dust destruction radius used in
\cite{2004MNRAS.351..607W}. The initial disc scaleheight at 100\,AU,
$h(100)$, was set to 25\,AU. As the simulation used vertical
hydrostatic equilibrium and dust sublimation, both the disc
scaleheight and inner edge location then evolved in the systems
dependent on the input parameters. In this section we discuss the
structure of the discs in terms of these two generated
characteristics, i.e the disc scaleheight and inner edge location.

@@

As discussed in \cite{2004MNRAS.351..607W} discs around CTTS stars
have scaleheights at 100 AU of between $h$(100)$=$7 to 20 AU, whereas
for BDD systems, $h$(100)=20 to 60 AU (for 0.08 and 0.01 $M_{\odot}$
respectively). Our models adopt the same formalism for solving
hydrostatic equilibrium and therefore, at negligible accretion rates,
our disc scaleheights are comparable. As the accretion rate increases
the flux levels of the central star increase and lead to heating of
the disc which in turn leads to vertical expansion. For
our models we found that levels of vertical flaring increased
marginally with accretion rate. Significant differences, more than
$>$5 AU increase in $h(50)$, in the vertical structure were not
apparent until the high accretion rates of \logmdot = $-7$ and
$-6$. Figures \ref{flare_-12_183} and \ref{flare_-7_193} show the
density structure ($log({\rho})$) in the disc from radial distances of
0 to 50 AU for example systems ($M_*=0.04M_{\odot}$, Age=1 Myrs,
$\tau$=5 and areal coverage=10\%), with accretion rates of \logmdot =
$-12$ and $-7$, respectively.

@@

\subsubsection{Inner edge}
\label{inner_edge}

@@

\subsubsection{Accretion dominance}
\label{acc_dominance}

@@

\subsubsection{Flaring and obscuration}
\label{flare_obscure}
